
\chapter{Tutorial}
\label{chapt:tutorial}

Welcome to the Ada programming language! The purpose of this tutorial is to give you an overview
of Ada so that you can start writing Ada programs quickly. This tutorial does not attempt to
cover the entire language. Ada is very large, so complete coverage of all its features would
take many more pages than are contained in this document. However, it is my hope that after
reading this tutorial you will have a good sense of what Ada is like, appreciate some of its
nicer features, and feel interested in learning more
\cite{Barnes2014,McCormick2011,Ben-Ari2009,Burns2007,Dale2007}.

This tutorial assumes you are already familiar with one or more languages in the C family: C,
C++, Java, C\#, or something similar. It is not my intention to teach you how to program. I
assume you already understand concepts such as loops, data types, functions, and so forth.
Instead, this tutorial describes how to use these things in Ada. In cases where Ada provides
features that might be unfamiliar to you (such as subtypes, discriminated types, and tasking) I
will discuss those features a bit more comprehensively.

Ada is a powerful language designed to address the following issues:

\begin{itemize}
\item The development of very large programs by multiple, loosely connected teams. The language
  has features to help manage many program components, and to help ensure those components are
  used consistently.

\item The development of long-lived programs that spend most of their time in the maintenance
  phase of the software life cycle. The language is designed to promote the readability of
  programs. You may find Ada code to be rather verbose and tedious to write. However, that extra
  work pays off later by making the code clearer and easier to read when bugs must be fixed or
  enhancements written.

\item The development of robust programs where correctness, security, and reliability are
  priorities. The language has features designed to make programming safer and less error-prone.
  Some of these features involve extra run time checking and thus entail a performance penalty.
  However, Ada's design is such that the performance penalty is normally small.

\item The development of embedded systems where low-level hardware control, multiple concurrent
  tasks, and real time requirements are common. The language has features designed to support
  these things while still retaining as much safety as feasible.
\end{itemize}

\input{Hello-Ada}
\input{Control-Structures}
\input{Types-and-Subtypes}
\input{Subprograms}
\input{Arrays-and-Records}
\input{Packages}
\input{Abstract-Data-Types}
\input{Strings}
\input{Exceptions}
\input{Discriminated-Types}
\input{Generics}
\input{Access-Types}
\input{Command-Line-Arguments}
\input{Object-Oriented-Programming}
\input{Tasking}
\input{Container-Library}
\input{Low-Level-Programming}
